Insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry

### 2019-04-03 analysis/paper/paper.Rmd
make multiple outcome plotting fun

---
title: "Insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik^1^
  - Sabine Dölle-Bierke^1^
  - Franziska Ruëff^2^ 
  - Claudia Pföhler^3^
  - Kathrin Scherer Hofmeier^4^
  - Margitta Worm^1^
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

```{r insectstime, fig.cap="Proportion of anaphylaxis cases elicited by specific insects according to the month in which the reaction occured. Less frequent insects were groupeped together as 'other'. The lowest panel shows the proportion of insect elicited cases to ceses elicited by other triggers."}
plot.proportions(rdb, "d_centres_country","q_340_insects",20)
```

@@

```{r insectsgeo, fig.cap="Proportion of anaphylaxis cases elicited by specific insects in reporting countries. Less frequent insects were groupeped together as 'other'. Countries which reported less than 20 cases of anaphylaxis due to insect venom were not illustrated in this figure."}
plot.proportions(rdb, "d_centres_country","q_340_insects",20)
```

@@

@@

@@

@@

```{R adrenalineuse, fig.cap = "The proportion of insect venom elicited anaphylaxis cases that were treated with adrenaline in comparison to other elicitors."}
F2$plot2

>>>>

---
title: "Insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik^1^
  - Sabine Dölle-Bierke^1^
  - Franziska Ruëff^2^ 
  - Claudia Pföhler^3^
  - Kathrin Scherer Hofmeier^4^
  - Margitta Worm^1^
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    #bookdown::html_document2:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

```{r insectstime, fig.cap="Proportion of anaphylaxis cases elicited by specific insects according to the month in which the reaction occured. Less frequent insects were groupeped together as 'other'. The lowest panel shows the proportion of insect elicited cases to ceses elicited by other triggers.", fig.height=6,fig.width=4}
#### plot MOR#####
plot_MOR <- gridExtra::grid.arrange(
  #cowplot::plot_grid(
  rdbp %>%
  select(b_reactiondate,grouping,q_340_insects,d_centres_country) %>%
  mutate(MOR = substr(b_reactiondate,4,5)) %>%
  filter(!is.na(q_340_insects),MOR!="00") %>%
  ggplot(aes(MOR,fill=q_340_insects))+
  geom_bar(position = "fill")+
    theme_classic()+
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none"
        )+
  ylab("Proportion"),
rdbp %>%
  select(b_reactiondate,grouping,q_340_insects,d_centres_country) %>%
  mutate(MOR = substr(b_reactiondate,4,5)) %>%
  filter(!is.na(q_340_insects),MOR!="00") %>%
  ggplot(aes(MOR,fill=q_340_insects))+
  geom_bar()+
  theme_classic()+
  labs(fill="Insect",x = "Month of the year")+
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = c(0.01,.98),
        legend.justification = c(0,1)),
rdbp %>%
  select(b_reactiondate,grouping,d_elicitor_gr5,d_centres_country) %>%
  mutate(MOR = substr(b_reactiondate,4,5),
         d_elicitor_gr5 = relevel(d_elicitor_gr5, "insects")) %>%
  filter(!is.na(grouping),MOR!="00") %>%
  group_by(MOR,grouping) %>%
  summarize(n = n()) %>%
  group_by(MOR) %>%
  summarise(prop = n[1]/sum(n)) %>%
  ggplot(aes(MOR,prop))+
  geom_bar(stat="identity")+
  theme_classic()+
  labs(x = "Month of the year",y = "Fraction of insect elicited ANA"),
  #theme(legend.position = c(0.01,.98),
  #      legend.justification = c(0,1)),
heights = c(0.4,1,0.5),
ncol = 1
)
```

@@

```{r insectsgeo, fig.cap="Proportion of anaphylaxis cases elicited by specific insects in reporting countries. Less frequent insects were groupeped together as 'other'. Countries which reported less than 20 cases of anaphylaxis due to insect venom were not illustrated in this figure."}
plot.proportions(rdb, "d_centres_country","q_340_insects",5)+
  labs(x = "Country",y="Proportion of IVA cases",fill = "Species")
```

@@

Cases of IVA were more requent in oder patients. Children suffered mainly form food elicited anaphylaxis. Therefore we adjusted the anlysis apropriately for age.

@@

### ACE-I
Ruef wrote in 2009 that the use of ACE I increases the risk of a severe reaction.

```{r}
data$q_423_ace <- yesonly(data$q_423_ace)

wilcox.test(q_116_VAS_v7~q_423_ace,data = data)
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_423_ace,q_116_VAS_v7))+
  geom_violin()

data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
group_by(q_423_ace) %>% 
  summarise(medianVAS=median(q_116_VAS_v7,na.rm=T),
            IQRVAS = IQR(q_116_VAS_v7,na.rm=T))

wilcox.test(ANAscore~q_423_ace,data = data %>% 
  filter(d_elicitor_gr5=="insects") 
)

data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_423_ace,ANAscore))+
  geom_violin()
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
 group_by(q_423_ace) %>% 
  summarise(medianANAscore=median(ANAscore,na.rm=T),
            IQRANAscore = IQR(ANAscore,na.rm=T))

cowplot::plot_grid(
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_423_ace,q_116_VAS_v7))+
  geom_boxplot(),
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_423_ace,ANAscore))+
  geom_boxplot()
)

### Beta blockers 
cardioselectivity??????
```{r}
data$q_423_beta <- yesonly(data$q_423_beta)

wilcox.test(q_116_VAS_v7~q_423_beta,data = data)
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_423_beta,q_116_VAS_v7))+
  geom_violin()

data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
group_by(q_423_beta) %>% 
  summarise(medianVAS=median(q_116_VAS_v7,na.rm=T),
            IQRVAS = IQR(q_116_VAS_v7,na.rm=T))

wilcox.test(ANAscore~q_423_beta,data = data %>% 
  filter(d_elicitor_gr5=="insects") 
)

data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_423_beta,ANAscore))+
  geom_violin()
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
 group_by(q_423_beta) %>% 
  summarise(medianANAscore=median(ANAscore,na.rm=T),
            IQRANAscore = IQR(ANAscore,na.rm=T))

cowplot::plot_grid(
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_423_beta,q_116_VAS_v7))+
  geom_boxplot(),
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_423_beta,ANAscore))+
  geom_boxplot()
)

### mast cell burden - mastocytosis

```{r}
data$q_410_masto_cur %<>%  yesonly()

wilcox.test(q_116_VAS_v7~q_410_masto_cur,data = data)
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_410_masto_cur,q_116_VAS_v7))+
  geom_violin()

data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
group_by(q_410_masto_cur) %>% 
  summarise(medianVAS=median(q_116_VAS_v7,na.rm=T),
            IQRVAS = IQR(q_116_VAS_v7,na.rm=T))

wilcox.test(ANAscore~q_410_masto_cur,data = data %>% 
  filter(d_elicitor_gr5=="insects") 
)

data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_410_masto_cur,ANAscore))+
  geom_violin()
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
 group_by(q_410_masto_cur) %>% 
  summarise(medianANAscore=median(ANAscore,na.rm=T),
            IQRANAscore = IQR(ANAscore,na.rm=T))

cowplot::plot_grid(
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_410_masto_cur,q_116_VAS_v7))+
  geom_boxplot(),
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_410_masto_cur,ANAscore))+
  geom_boxplot()
)

```{r}
cowplot::plot_grid(
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_212_tryptase_value_v5,q_116_VAS_v7))+
  geom_point()+
  scale_x_log10(),
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
ggplot(aes(q_212_tryptase_value_v5,ANAscore))+
  geom_point()+
  scale_x_log10()
)

cowplot::plot_grid(
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
  mutate(typ_cat=ifelse(q_212_tryptase_value_v5<8,
                  "low",
                  "high")) %>% 
  filter(!is.na(typ_cat)) %>% 
  ggplot(aes(typ_cat,q_116_VAS_v7))+
  geom_boxplot(),
data %>% 
  filter(d_elicitor_gr5=="insects") %>% 
  mutate(typ_cat=ifelse(q_212_tryptase_value_v5<8,
                  "low",
                  "high")) %>% 
  filter(!is.na(typ_cat)) %>% 
  ggplot(aes(typ_cat,ANAscore))+
  geom_boxplot(),
data %>% 
  filter(d_elicitor_gr5=="insects",
         !is.na(d_severity_rm)) %>% 
  mutate(typ_cat=ifelse(q_212_tryptase_value_v5<8,
                  "low",
                  "high")) %>% 
  group_by(d_severity_rm,typ_cat) %>% 
  filter(!is.na(typ_cat)) %>% 
    summarize(n = n()) %>% 
  ggplot(aes(typ_cat,n,fill=d_severity_rm))+
  geom_bar(stat="identity",position= "fill"),
data %>% 
  filter(d_elicitor_gr5!="insects") %>% 
  mutate(typ_cat=ifelse(q_212_tryptase_value_v5<8,
                  "low",
                  "high")) %>% 
  filter(!is.na(typ_cat)) %>% 
  ggplot(aes(typ_cat,q_116_VAS_v7))+
  geom_boxplot(),
data %>% 
  filter(d_elicitor_gr5!="insects") %>% 
  mutate(typ_cat=ifelse(q_212_tryptase_value_v5<8,
                  "low",
                  "high")) %>% 
  filter(!is.na(typ_cat)) %>% 
  ggplot(aes(typ_cat,ANAscore))+
  geom_boxplot(),
data %>% 
  filter(d_elicitor_gr5!="insects",
         !is.na(d_severity_rm)) %>% 
  mutate(typ_cat=ifelse(q_212_tryptase_value_v5<8,
                  "low",
                  "high")) %>% 
  group_by(d_severity_rm,typ_cat) %>% 
  filter(!is.na(typ_cat)) %>% 
    summarize(n = n()) %>% 
  ggplot(aes(typ_cat,n,fill=d_severity_rm))+
  geom_bar(stat="identity",position= "fill")
)

Here put the mastocytosis and baseline serum tryptase as the evaluation of severe risk
Here it is crucial to compare to even these reactions that had nearly no reactions

### Model the risk of severity of a severe reaction in IVA

USe the model from our previous paper
```
```

@@

### Do we know how mnay beekeepers were in our dataset?
We were unable to estimate the fraction of cases that were elicited in bee-keepers as this information is not asked in the questionnaire.

### Hypothesis: Wasp stings are more severe than bee stings
We previously saw that refractory anaphylaxis cases come from bees. Here we see no difference in the severity of cases elicited by bees and wasps (in the RM).

```{r}
ana_outcome_matched <- function(data = data4,
         outcome = "d_severity_rmr",
         matching_var = "q_340_insects",
         confounders_cat = c("b_sex", "cc_cardio"),
         confounders_int = c("d_age"),
         selected_predictors = c(1:4)
         ){
  possible_outcomes <- c("d_severity_rm",
                         "d_severity_rmr",
                         "severity_brown",
                         "q_116_VAS_v7",
                         "ANAscore")
  predictors <- c(matching_var, confounders_cat,confounders_int)
  df <- data %>% # first get the data frame
   filter(!!as.name(matching_var) %in% c("yellow jacket", "bee")) %>% 
   select(!!!matching_var,
          !!!outcome,
          !!!confounders_cat,
          !!!confounders_int
          ) %>%
          {.[complete.cases(.),]}
  ### we need to match the cases now to the first level of theoutcome variable
  ref_level <- levels(df[,matching_var] %>% droplevels())
  matched <- df %>% 
    mutate(match = ifelse(!!as.name(matching_var)==ref_level[1],T,F),
           out=ifelse(!!as.name(outcome)=="severe",1,0)) %>% 
    matchit(formula = as.formula(
      paste("match ~ ",paste0(confounders_cat),"+", paste0(confounders_int,collapse = " + ")))) %>% 
    match.data() 
  ## IFELSE
  if(outcome %in% possible_outcomes[c(2,3)]){
    plot <- matched %>% 
      ggplot(aes(!!as.name(matching_var),
               fill = !!as.name(outcome)))+
      geom_bar()+
      facet_grid(reformulate(confounders_cat[1],confounders_cat[2]))

    fit <- glm(as.formula(paste0(
      "out ~ ",paste0(predictors[selected_predictors], collapse = " + ")
    )),data= matched,family= "binomial")
  } else if (outcome %in% possible_outcomes[c(1,4,5)]){
     plot <- matched %>% 
      ggplot(aes(!!as.name(matching_var),
               y = !!as.name(outcome)))+
      geom_boxplot()+
      facet_grid(reformulate(confounders_cat[1],confounders_cat[2]))

    fit <- glm(as.formula(paste0(
      outcome, " ~ ", paste0(predictors[selected_predictors], collapse = " + "))),
             data = matched)
  }

  return(list(mtched_df = matched,
              ref_level = ref_level,
              fit = fit,
              plot = plot)) 
}

possible_outcomes <- c("d_severity_rm",
                         "d_severity_rmr",
                         "severity_brown",
                         "q_116_VAS_v7",
                         "ANAscore")

output_severity <- lapply(possible_outcomes,
                         function(x){
                         ana_outcome_matched(data = data4,
                            outcome = x,
                            matching_var = "q_340_insects",
                            confounders_cat = c("b_sex", "cc_cardio"),
                            confounders_int = c("d_age"),
                            selected_predictors = c(1:4))
                           }
)

do_cof_tab <- function(x){
  output_severity[[x]]$fit %>% summary() %>% {.$coefficients}%>% tidy() %>% as.matrix()
}

temp <- lapply(1:5,do_cof_tab) %>% do.call(what = rbind) %>%
                          as_tibble() %>%
  {mutate(.,
         Estimate = as.numeric(Estimate),
         Pr...t.. = as.numeric(Pr...t..)
  )} %>%
          {cbind(.,outcome = lapply(possible_outcomes,
                          rep,
                          nrow(.)/5) %>% unlist)
          } %>% 
           filter(.rownames!="(Intercept)")
gridExtra::grid.arrange(output_severity[[2]]$plot,
                        output_severity[[3]]$plot,
                        output_severity[[1]]$plot,
                        output_severity[[4]]$plot,
                        output_severity[[5]]$plot,
                        temp %>% 
  ggplot(aes(.rownames,Pr...t.., color=outcome))+
  geom_segment(x=0.5,xend=4.5,y=log10(0.05),yend=log10(0.05))+
  scale_y_log10()+
  geom_jitter()+
  theme(axis.text.x = element_text(angle=40, hjust=1))
  )#+
  #geom_line(aes(group=outcome))
```

Based on the above graphs of patients matched according to sex and age we cannot say that the severity of their reaction differ between yellow-jackets and bees in any of the models except ANAscore - where yellow jacket were linked to  a slightly more severe reaction.

But in the ANAscore variable it shows up to be significant ! Yellowjackets stings are more severe than the bee stings. This is however very hard to compare to refractory cases. Maybe we can use the refractory case identifier to say in which group we see more refractory cases?

### Increased BTSC develops in the normal range <8

```{r}
output_severity <- lapply(possible_outcomes,
                         function(x){
                         ana_outcome_matched(data = data4,
                            outcome = x,
                            matching_var = "q_340_insects",
                            confounders_cat = c("tryp_cat", "cc_cardio"),
                            confounders_int = c("d_age"),
                            selected_predictors = c(1,2,4))
                           }
)

do_cof_tab <- function(x){
  output_severity[[x]]$fit %>% summary() %>% {.$coefficients}%>% tidy() %>% as.matrix()
}

temp <- lapply(1:5,do_cof_tab) %>% do.call(what = rbind) %>%
                          as_tibble() %>%
  {mutate(.,
         Estimate = as.numeric(Estimate),
         Pr...t.. = as.numeric(Pr...t..)
  )} %>%
          {cbind(.,outcome = lapply(possible_outcomes,
                          rep,
                          nrow(.)/5) %>% unlist)
          } %>% 
           filter(.rownames!="(Intercept)")
gridExtra::grid.arrange(output_severity[[2]]$plot,
                        output_severity[[3]]$plot,
                        output_severity[[1]]$plot,
                        output_severity[[4]]$plot,
                        output_severity[[5]]$plot,
                        temp %>% 
  ggplot(aes(.rownames,Pr...t.., color=outcome))+
  geom_segment(x=0.5,xend=4.5,y=log10(0.05),yend=log10(0.05))+
  scale_y_log10()+
  geom_jitter()+
  theme(axis.text.x = element_text(angle=40, hjust=1))
  )
```

Based on the above we see that the lower tryptase significantly decreased the severity of

### Mastocytosis

### Older Age

### ACE - I

### previous reactions (severe)

### Male sex

### Beta blockers and CARDIOSELECTIVE BLOCKERS

### Evaluate the pediatric population separately

### Number of SAR during VIT compare to other hypos
#### Compare bee sit to wasp sit for AEs
#### correlate the tryptase value with SIT types and severity
### 
they used generalized aditive models (package GAM?)
Also ROC curves for confirmation

@@

```{R adrenalineuse, fig.cap = "The proportion of insect venom elicited anaphylaxis cases that were treated with adrenaline in comparison to other elicitors."}
F2$plot2
```
